A Triangle Mesh Approach to Painting

### 2014-04-20 template.tex
Draft for edits

\begin{abstract}

@@

All computer painting programs must store data through rasterization or vectorization. 
However, traditional computer painting programs take inputs of the same type as the stored data. The 
extremely popular Adobe programs Photoshop and Illustrator exemplify this. The former
allows a user to paint how he would on a physical medium. By letting the mouse act as
a brush, a user may color individual pixels on the screen. The latter lets a user paint
using geometric primitives such as lines and B\'{e}zier curves. The program then stores
these geometric primitives in order to render the drawing. Although using vector graphics
is likely less intuitive, it has some distinct advantages. First, there is no resolution
limit. One can zoom into an image as much as he would like and still see crisp edges.
Second, certain editing operations may be easier to perform on vector graphics.

In this paper we introduce TrianglePainter which  
uses a slightly different representation than traditional raster or vector graphics.
We instead use triangle meshes to store the data of the painting.
Since triangles are a geometric primitive used in vector graphics our approach achieves the
same advantages traditional vector graphics have. However, our representation allows the
user to paint with vector graphics without worrying about the underlying implementation.

@@

Furthermore, our representation allows painting at any scale. A user can make large strokes
while zoomed out and then zoom in to make fine details. All of this can happen without
loss of quality since the data is stored using triangles, not in pixels.

To do the transformation we use a combination of rasterization and marching squares to
determine the contour of the stroke. From the contour we can create a Delaunay triangulation
to represent the stroke. By then introducing a merging process we can take the triangulation
of the new stroke and combine it with the triangulation of the existing canvas. This yields
a new canvas with all the strokes combined.

TrianglePainter reduces the barriers in creating vector graphics. This in turn allows 
artists to focus on the painting and not the way in which they paint, while still 
getting the editing benefits of vector graphics.

@@

Other work has proposed novel vector graphic data structures. For example \cite{Frisken:2000:ASD:344779.344899}
defines Adaptively Sampled Distance Fields (ADF). ADFs specify a signed distance function to a surface.
to render the shape requires sampling the function to determine whether or not a pixel is on the surface.
A quadtree accompanies the distance function to specify where higher sampling rates are necessary. \cite{Bremer:2001:VCM}
further describes how ADFs can be created and used. However, it is unclear how ADFs might be used in a general
painting program where strokes with solid colors and gradients are composited on top of one another.

Another series of works have created entirely new ways of painting. Diffusion curves as specified in \cite{Orzan:2008:DCV:1360612.1360691}
describe a new painting technique where the artists specify edges and colors for those edges. Their system
then solves a Poisson equation to diffuse the colors between boundary conditions specified at the edges.
The algorithm produces some truly beautiful drawing. Similarly \cite{McCann:2008:RGP:1360612.1360692}
lets artists paint in the gradient domain. Both of these approaches yield fantastic works of art, yet
they are really new mediums for artists to explore rather than a vectorized version of conventional
painting techniques.

Work has also been done in multi-resolution painting using raster graphics. \cite{Berman:1994:MPC:192161.192181}
stores image information in a quadtree so that different parts of the image can have different levels of detail.
This means the image effectively has an infinite level of detail, yet since it is stored in raster form
there are still limitations. For example a stroke painted at coarse detail will still look blurry when
zoomed in. The strokes are limited to the resolution they are painted at. \cite{Carr:2004:PD:1186562.1015809}
also allows the development of multi-resolution images using rasters, and as such suffers from the
same limitations. Another multi-resolution approach by \cite{Perlin:1995:LPP:218380.218437} uses
procedural textures to avoid any fundamental limit on resolution. Although useful for textures,
generating procedural textures is difficult and does not help artists easily create.

@@

\section{The Canvas Model}
The data necessary to render the painting resides in the canvas. However, the canvas contains more than just
a triangle mesh and understanding the underlying data structures is important for 
understanding the rest of the algorithm.

@@

\subsection{Boundary Edges}
Although the triangles may change over time, they must abide by constraints known as
boundary edges. A boundary edge connects two points and specifies that no triangle
can cross it. Like points, boundary edges are persistent. So as triangles come and go they
may never cross a boundary edges.

\subsection{Results}

@@

\includegraphics[height=1.5in]{images/softstrokeboundaryedges}

@@

Here we consider how to convert a user's cursor motion into a triangle-mesh representation
of their stroke.

@@

Adam and Steve should we talk about how anti-aliasing + marching squares can achieve sub-pixel
resolution?

@@

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardrendered}
  \caption{A subfigure}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardpruned}
  \caption{A subfigure}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardmesh}
  \caption{A subfigure}
  \label{fig:sub2}
\end{subfigure}
\caption{A figure with two subfigures}
\label{fig:test}
\end{figure*}

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softrendered}
  \caption{A subfigure}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softpruned}
  \caption{A subfigure}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softmesh}
  \caption{A subfigure}
  \label{fig:sub2}
\end{subfigure}
\caption{A figure with two subfigures}
\label{fig:test}
\end{figure*}

@@

This representation needs a slightly different procedure. The user's mouse motion
is capture in two sets of polygons. The first set describes the outer soft stroke and the second set describes
the inner hard stroke. Next the outer stroke polygons are rendered in 50\% gray and the inner stroke
is rendered in black. This means the boundary of the outer stroke is the 25\% gray iso-contour and the
boundary of the inner stroke is the 75\% gray iso-contour.

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are fed into Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard parts
of the stroke.

\includegraphics[height=1.5in]{images/softstroke}
\includegraphics[height=1.5in]{images/hardvssoft}

\section{Compositing Strokes}
Once the stroke has been converted to triangles by the above procedure it is necessary to
composite the stroke onto the canvas. We first consider the canvas model and then how a
stroke can be composited onto the canvas.

@@

Every canvas begins with four points, one at each corner of the screen, and four boundaries, one
for each edge of the screen. These points are originally part of two big triangles which
divide the canvas.

@@

\includegraphics[height=1.5in]{images/canvastriangulation}

@@

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
points on the boundaries. For both boundaries travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points without color data. For both boundaries linearly interpolate the colors in the color arcs. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections.

\includegraphics[height=1.5in]{images/determiningcolors}

@@

\includegraphics[height=1.5in]{images/findingmodifiedpoints}

\section{Discussion and Results}
Our work with this algorithm has lead to observations in three main areas.

\subsection{Stability}
Something we've worked hard to achieve is program stability. Certain geometries are fatal
for the Triangle library. The current instability in the program is normally caused by
providing the library with such difficult geometries. To improve program stability would
require addressing geometries such as vertices that are too close together. This would
take more processing time and is something we have yet to explore. However, clearly this
problem must be solved before the program could be used formally.

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to extract a stroke from rasterization to triangle
representation on the canvas is slow. On a blank canvas strokes take on the order of
1 second to process. However, on a canvas with lots of geometry a stroke that covers
that geometry can take several seconds to render. Some of this time is due to poor
implementation of the algorithm. However, there are lots of calculations
to do with the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can get arbitrarily complicated, these
calculations can take an arbitrarily long. This brings us to the final point, scalability.

\subsection{Scalability}
No matter how good the implementation, the fact that paintings can get arbitrarily complicated
means there must be some way to simplify the mesh. Although we have yet to implement
such a simplification algorithm, we have a couple suggestions on how it might be done. In most situations where
a lot of strokes are composited on top of one another, most of the geometry is redundant. 
For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it. Furthermore, even with translucent or feathered strokes, enough
composited strokes may produce complex geometry that does not add much to the output.
Some of this geometry can be reduced without noticeable changes in the output. Such a
simplification algorithm must be implemented to ensure a painter can continue to paint
without the program slowing to an unusable state.

\subsection{Complexity}
We now examine the growth of triangles, points, and boundaries on the canvas.

@@

The complexity of the triangles increase slightly more than linearly since extra points and 
boundaries are made when boundaries intersect.

NOTES:

-Maybe do the same analysis with soft strokes

-Look at the time it takes to render these strokes

-Then show more robust drawings and the number of triangles, points, and boundaries

\section{Conclusion}

\section*{Acknowledgements}

>>>>

\begin{abstract}
Vector graphics have a compact representation, are easy to manipulate, and have infinite resolution.
Traditionally, vector graphics have required artists to master a new skill in order to effectively utilize
them. Recent work has converted raster graphics to vector graphics, allowed multi-resolution raster painting,
and created new ways to represent vector graphics. However, none of these have solved the problem of
letting an artist use standard and intuitive raster painting techniques to generate vector graphics.
Our algorithm, TrianglePainter, converts a user's strokes into a triangle mesh vector representation. 
TrianglePainter lets artists focus on the art, while it handles the vectorization in the background.

@@

All computer painting programs must store data through rasterization or vectorization. 
However, traditional computer painting programs take inputs of the same type as the stored data. 
For example, programs like Photoshop and GIMP
allow a user to paint how he would on a physical medium. By letting the mouse act as
a brush, a user may color individual pixels on the screen. By contrast, programs like 
Illustrator and Inkscape let a user paint
using geometric primitives such as lines and B\'{e}zier curves. The program then stores
these geometric primitives in order to render the drawing. Although using vector graphics
is likely less intuitive, it has some distinct advantages including a compact
representation, infinite resolution, and easier manipulation.

In this paper we introduce TrianglePainter which  
uses a slightly different representation than traditional raster or vector graphics.
TrianglePainter uses a triangle mesh to store the data of the painting.
Since triangles are a geometric primitive used in vector graphics this approach achieves the
same advantages traditional vector graphics have. However, the representation allows the
user to paint with vector graphics without dealing directly with the underlying implementation.

@@

Furthermore, this representation allows painting at any scale. A user can make large strokes
while zoomed out and then zoom in to make fine details. All of this can happen without
loss of quality since the data is stored using triangles, not in pixels.

To transform a user's mouse inputs to a triangle mesh
TrianglePainter uses a combination of rasterization and marching squares to
find stroke contours. From the contour it performs a Delaunay triangulation
and then merges this triangulation
of the new stroke with the triangle mesh of the existing canvas. This process yields
a new canvas with the combined strokes.

TrianglePainter reduces the barrier to creating vector graphics. This in turn allows 
artists to focus on the painting and not the way in which they paint, while still 
getting the benefits of vector graphics.

@@

Other work has proposed novel vector graphic data structures. For example \cite{Frisken:2000:ASD:344779.344899}
defines Adaptively Sampled Distance Fields (ADF). ADFs specify a signed distance function to a surface.
Rendering the shape requires sampling the function to determine whether or not a pixel is on the surface.
A quadtree accompanies the distance function to specify where higher sampling rates are necessary. \cite{Bremer:2001:VCM}
further describes how ADFs can be created and used. However, it is unclear how ADFs might be used in a general
painting program where strokes with solid colors and gradients are composited on top of one another.

Another series of works have created entirely new ways of painting. Diffusion curves as specified in \cite{Orzan:2008:DCV:1360612.1360691}
describe a new painting technique where the artists specify edges and colors for those edges. Their system
then solves a Poisson equation to diffuse the colors between boundary conditions specified at the edges.
Similarly \cite{McCann:2008:RGP:1360612.1360692}
lets artists paint in the gradient domain. Both of these approaches yield fantastic works of art, yet
they are really new mediums for artists to explore rather than a vectorized version of conventional
painting techniques.

Work has also been done in multi-resolution painting using raster graphics. \cite{Berman:1994:MPC:192161.192181}
stores image information in a quadtree so that different parts of the image can have different levels of detail.
This means the image effectively has an infinite level of detail, yet since it is stored in raster form
there are still limitations. For example a stroke painted at coarse detail will still look blurry when
zoomed in. The strokes are limited to the resolution they are painted at. \cite{Carr:2004:PD:1186562.1015809}
also allows the development of multi-resolution images using rasters, and as such suffers from the
same limitations. Another multi-resolution approach by \cite{Perlin:1995:LPP:218380.218437} uses
procedural textures to avoid any fundamental limit on resolution. Although useful,
generating procedural textures is difficult and does not help most artists easily create.

@@

\section{The Canvas Model}
The data necessary to render the painting resides in the canvas. However, the canvas contains more than just
a triangle mesh, and understanding the underlying data structures is important for 
understanding the rest of the algorithm.

@@

\subsection{Boundary Edges}
Although the triangles may change over time, they must abide by constraints known as
boundary edges. A boundary edge connects two points and specifies that no triangle
can cross it. Like points, boundary edges are persistent. So as new triangles are created they
may never cross a boundary edges.

@@

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/softstrokeboundaryedges}
    \caption{Place Holder.}
\end{figure}

@@

Here we consider how to convert a user's cursor motion into a triangle-mesh representation
of their stroke. This is done in two steps. First, the stroke is rasterized and the
contour is extracted from the raster. Second, the contour is converted into points and
boundaries and triangulated.

@@

% Adam and Steve should we talk about how anti-aliasing + marching squares can achieve sub-pixel
% resolution?

@@

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardrendered}
  \caption{The rasterized stroke.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardpruned}
  \caption{The contour of the stroke.}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/hardmesh}
  \caption{The triangulation of the contour}
  \label{fig:sub2}
\end{subfigure}
\caption{Triangulating a hard stroke.}
\label{fig:test}
\end{figure*}

\begin{figure*}
\centering
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softrendered}
  \caption{The rasterized stroke.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softpruned}
  \caption{The contour of the stroke.}
  \label{fig:sub2}
\end{subfigure}
\begin{subfigure}[b]{0.3\textwidth}
  \centering
  \includegraphics[width=.9\textwidth]{images/stroke_triangulation/softmesh}
  \caption{The triangulation of the contour}
  \label{fig:sub2}
\end{subfigure}
\caption{Triangulating a soft stroke.}
\label{fig:test}
\end{figure*}

@@

This representation needs a slightly different procedure. The user's mouse motion
is captured in two sets of polygons. The first set describes the outer soft stroke and the second set describes
the inner hard stroke. Next the outer stroke polygons are rendered in 50\% gray and the inner stroke
is rendered in black. This means the boundary of the outer stroke is the 25\% gray iso-contour and the
boundary of the inner stroke is the 75\% gray iso-contour.

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are fed into Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard parts
of the stroke. Points on the contour of the inner stroke will have color arcs with the stroke color
while the points on the contour of the outer stroke will have transparent color arcs. This gives
all triangles connecting the inner and outer strokes a linear gradient.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/softstroke}
        \includegraphics[height=1.5in]{images/hardvssoft}
    \caption{Place Holder.}
\end{figure}

\section{Compositing Strokes}
Once the stroke has been converted to triangles by the above procedure it is necessary to
composite the stroke onto the canvas. This is accomplished in two steps. First, the points
and boundaries from the triangulation of the stroke are added to the canvas and a new
triangulation is found. Second, the color arcs for the old and new points are updated.
We also discuss the challenge intersecting boundaries pose for the algorithm, as well
as a technique to reduce the number of triangles that are retriangulated on the canvas.

@@

Every canvas begins with a number of points, along the edge of the screen. These
points have boundaries connecting them. These points are originally part of a number of triangles which
divide the canvas.

@@

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/canvastriangulation}
    \caption{Place Holder.}
\end{figure}

@@

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
points on the boundaries. For both boundaries travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points without color data. For both boundaries the colors in the color arcs are linearly interpolated. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/determiningcolors}
    \caption{Place Holder.}
\end{figure}

@@

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/findingmodifiedpoints}
    \caption{Place Holder.}
\end{figure}

\section{Discussion and Results}
Here we first consider how the complexity of the algorithm grows both in time and geometry.
The results from this analysis and our experience with TrianglePainter have lead to observations in three main areas.

\subsection{Complexity}
To find the growth rates of the algorithm we performed two experiments. 
First, the program drew fifty random strokes without zooming.
Next, the program drew fifty random strokes with random zooming.
Geometry grows fastest when many strokes are composited on one another. When not zooming strokes
must cover the same area of canvas over and over. This
results in more geometry and therefore takes longer to render. The zooming case explores different
areas of the canvas and therefore results in less overlapping geometry. This situation should result
in less complex geometry and therefore take less time.

\begin{figure}
    \centering
        \includegraphics[width=0.49\textwidth]{graphs/numtriangles}
    \caption{The blue graph is the non-zooming case. The red graph is the zooming case.}
    \label{fig:numtriangles}
\end{figure}

As seen in figure~\ref{fig:numtriangles} the zooming case grows linearly while the non-zooming
appears to grow quadratically. This is expected since every new stroke has the chance to overlap
all of the old strokes. This means each new stroke does not just add a constant amount of geometry.
Due to intersections with previous strokes intersection points are added. This results in a linear
increase in the amount of geometry. This results
in a quadratic growth rate. However, in the zooming case each stroke rarely overlaps and therefore
intersection points are rarely added. Instead the geometry increases by a constant which leads to
a linear growth rate.

\begin{figure}
    \centering
        \includegraphics[width=0.49\textwidth]{graphs/cumulativetime}
    \caption{The blue graph is the non-zooming case. The red graph is the zooming case.}
    \label{fig:timing}
\end{figure}

Figure~\ref{fig:timing} gives a similar result for the cumulative time strokes take to render. Both
the zooming and non-zooming cases grow non-linearly over time. However, the non-zooming case takes
a lot longer for the reasons described. It's not unexpected that even the zooming case grows non-linearly
since even though the geometries involved aren't getting very complicated, the geometry of the entire
canvas is increasing linearly and must be considered even if the entire canvas is not retriangulated.

All this suggests that a simplification is of great importance.
The case that it is most likely to help is the non-zooming case with complex geometry. When lots of strokes
overlap there is usually an extra geometry that does not add much detail to the drawing. By removing
this extra geometry the algorithm should achieve a steady growth rate.

\subsection{Stability}
Something we've worked hard to achieve is program stability. Certain geometries are fatal
for the Triangle library. The current instability in the program is normally caused by
providing the library with such difficult geometries. To improve program stability would
require addressing geometries such as vertices that are too close together. This would
take more processing time and is something we have yet to explore.

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to take a stroke from rasterization to triangle
representation on the canvas is slow. On a blank canvas strokes take on the order of
1 second to process. However, on a canvas with lots of geometry a stroke that covers
that geometry can take several seconds to render. Some of this time is due to poor
implementation of the algorithm. However, there are lots of calculations
to do with the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can get arbitrarily complicated, these
calculations can take an arbitrarily long. This brings us to the final point, scalability.

\subsection{Simplification}
No matter how good the implementation, the fact that paintings can get arbitrarily complicated
means there must be some way to simplify the mesh. Although we have yet to implement
such a simplification algorithm, we have a couple suggestions on how it might be done. In most situations where
a lot of strokes are composited on top of one another, most of the geometry is redundant. 
For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it. Furthermore, even with translucent or feathered strokes, enough
composited strokes may produce complex geometry that does not add much to the output.
Some of this geometry can be reduced without noticeable changes in the output. Such a
simplification algorithm must be implemented to ensure a painter can continue to paint
without the program slowing to an unusable state.

@@

% The complexity of the triangles increase slightly more than linearly since extra points and 
% boundaries are made when boundaries intersect.

% NOTES:

% -Maybe do the same analysis with soft strokes

% -Look at the time it takes to render these strokes

% -Then show more robust drawings and the number of triangles, points, and boundaries

% -Talk about future work in the conclusion

% - Switch 6.3 and 6.4
\begin{figure*}
    \centering
        \includegraphics[width=0.49\textwidth]{images/facezoom}
        \includegraphics[width=0.49\textwidth]{images/facezoomtriangles}
    \caption{Artist time ~30 minutes. 9030 triangles.}
\end{figure*}

\section{Conclusion}
Vector graphics are useful for representing complex images with limited information
while also enabling easy editing and infinite resolution. Previous work has focused
on converting raster graphics to vector graphics, unique vector representations, and multi-scale
raster graphics. However, until now no system
has enabled artists to create an entire painting in vector graphics using standard brush techniques.

TrianglePainter enables a user to paint like they would in a raster graphics program 
to create vector graphics. The core of TrianglePainter is a triangle mesh representation of the image.
We have created an algorithm to convert a user's mouse motions into a triangle mesh, and then composite
that mesh on the preexisting mesh containing the previous strokes.

The current implementation allows artists to create minimal works. Future work should focus on
implementing a simplification algorithm to allow the complexity of the drawings to grow. Other
improvements might focus on creating interesting transforms, or other operations that utilize
the underlying triangle mesh structure to make the artists' job easier.

\section*{Acknowledgments}
